---
title: "Galaxy Tool Suites"
lightbox: 
    match: auto
    effect: zoom
    desc-position: right
execute:
    eval: true
    warning: false
---

## Load `libraries` 

```{r message=FALSE}
# install.packages("data.table")
library(data.table)
# install.packages("stringr")
library(stringr)

# install.packages(c("ggplot2", "ggrepel", "ggtext", "ggh4x"))
library(ggplot2)
library(ggrepel)
library(ggtext)
library(ggh4x)

# install.packages(c("extrafont", "paletteer", "colorspace"))
library(extrafont)
library(paletteer)
library(colorspace)
```

## Load input dataset 

```{r loaddata, include = FALSE}

## THIS SECTION MIGHT THROUGH AN ERROR
## THERE IS A BUG UNFORTUNATELY WITH fread FUNCTION
## RUN ON CONSOLE IN THIS CASE

tools_url <- "https://raw.githubusercontent.com/usegalaxy-eu/microgalaxy_paper_2025/refs/heads/main/docs/supplementary/supplementary_table_2.tsv"

tools_dt <- tools_url |> fread()
```

## Tool Availability graph: Heatmap 

### Extract availability information

```{r}
index <- tools_dt |> colnames() |> str_subset("Number of tools on")

availability <- tools_dt[, c("Suite ID", "EDAM reduced topics", index), with = FALSE] |> unique()

colnames(availability) <- availability |> 
    colnames() |> 
    str_remove_all("Number\\ of\\ tools\\ on") |> 
    str_squish()
```

### Filter out duplicates

```{r}
index <- which(availability$`Suite ID` == "srst2" & availability$`EDAM reduced topics` == "")

availability <- availability[-index]
```

### Hierarchical clustering

```{r}
mm <- availability[, -c(1, 2)] |> as.matrix(rownames = availability$`Suite ID`)

mm_c <- mm |> dist(method = "manhattan") |> hclust(method = "ward.D2")
mm_r <- t(mm) |> dist(method = "manhattan") |> hclust(method = "ward.D2")
```

### Construct plotting data

```{r}
d <- availability |> melt(id.vars = c("Suite ID", "EDAM reduced topics"), variable.factor = FALSE, value.factor = FALSE)
d <- d[which(value >= 1)]

d$`Suite ID` <- d$`Suite ID` |> factor(levels = mm_c$labels[mm_c$order |> rev()])
d$variable   <- d$variable |> factor(levels = mm_r$labels[mm_r$order |> rev()])

d$fct <- ifelse(d$variable |> str_detect("UseGalaxy"), "UseGalaxy", "vOther")

index <- d[which(fct == "UseGalaxy")][[1]] |> unique()

p  <- d[which(`Suite ID` %in% index)]
p2 <- p |> tidyr::separate_rows("EDAM reduced topics", sep = ",") |> setDT()

p2$`EDAM reduced topics` <- p2$`EDAM reduced topics` |> str_squish()
p2$`EDAM reduced topics` <- p2$`EDAM reduced topics` |> str_wrap(width = 10)

t <- p2[, by = "EDAM reduced topics", .(N = `Suite ID` |> unique() |> length())]
t <- t[order(-N)]
t <- t[which(`EDAM reduced topics` != ""), head(.SD, 5)]

p2$edam_clean <- ifelse(p2$`EDAM reduced topics` %in% t$`EDAM reduced topics`, p2$`EDAM reduced topics`, "Other")
p2$edam_clean <- p2$edam_clean |> factor(levels = c(t$`EDAM reduced topics`, "Other"))
```

### Heatmap: Top 5 EDAM Operations

```{r}
c_1 <- p2[which(edam_clean != "Other")] |> 
    
    ggplot(aes(variable, `Suite ID`)) + 
    
    geom_tile(aes(fill = value), color = "grey") + 
    
    scale_fill_stepsn(
        colors = c('#00429d', '#5681b9', '#93c4d2', '#ffffe0', '#ffa59e', '#dd4c65', '#93003a'),
        guide = guide_colorsteps(barwidth = unit(14, "lines"), barheight = unit(.35, "lines")),
        breaks = c(2, 4, 8, 16, 32, 64, 128),
        transform = "log2"
    ) +
    
    facet_grid(cols = vars(fct), rows = vars(edam_clean), scales = "free", space = "free") +
    
    scale_y_discrete(expand = c(0, 0)) +
    scale_x_discrete(expand = c(0, 0)) +
    
    theme_minimal(base_family = "Calibri") +
    
    theme(
        legend.position = "bottom",
        legend.title.position = "top",
        
        axis.text.x = element_text(angle = 45, hjust = 1),
        axis.text.y = element_text(size = 6),
        
        axis.title.x = element_markdown(),
        
        panel.border = element_rect(fill = NA, color = "grey25"),
        axis.ticks.x = element_line(lineend = "round", color = "grey25"),
        
        strip.text.x = element_blank(),
        strip.text.y = element_text(face = "bold", angle = 0, hjust = 0),
        
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(linetype = "dashed", lineend = "round", linewidth = .25),
        
        plot.margin = margin(10, 10, 10, 10)
    ) +
    
    labs(y = "Galaxy Tool Suites", x = "Availability of Tool Suites Across **Servers**", fill = "No. of Tools")
```

## EDAM operation: Scatter plot 

### Split EDAM operation per tool

```{r}
df2 = tools_dt |> tidyr::separate_rows("EDAM reduced operations", sep = ",") |> setDT()

xvar <- "Suite runs (last 5 years) on main servers"
yvar <- "Suite users (last 5 years) on main servers"
    
df2$runs  <- df2[[ xvar ]]
df2$users <- df2[[ yvar ]]

```

### Remove empty tools

```{r}
df2 = df2[which(runs != 0 & users != 0)]
```

### Clean `EDAM operation (no superclasses)` column

```{r}
df2$`EDAM reduced operations` = df2$`EDAM reduced operations` |> str_squish()
df2$`EDAM reduced operations` = ifelse(df2$`EDAM reduced operations` == "", "No Operation", df2$`EDAM reduced operations`)
```

### Find most top EDAM operations

```{r}
st = df2[, by = `EDAM reduced operations`, .(N = `Suite ID` |> unique() |> length())]
st = st[order(-N)]
st = st[which(`EDAM reduced operations` != "No Operation")]

df2$`EDAM reduced operations` <- df2$`EDAM reduced operations` |> factor(levels = c(st$`EDAM reduced operations`, "No Operation"))

df2 <- df2[order(`Suite ID`, -`EDAM reduced operations`)]
df2 <- df2[, by = `Suite ID`, head(.SD, 1)]
```

### Keep only necessary columns

```{r}
df2 = df2[, c("Suite ID", "runs", "users", "EDAM reduced operations")] |> unique()
```

### Define factor levels of cluster column

```{r}
st <- st[1:12]

df2$cluster <- ifelse(df2$`EDAM reduced operations` %in% st$`EDAM reduced operations`, df2$`EDAM reduced operations` |> as.character(), "Other")
df2$cluster <- df2$cluster |> factor(levels = c(st$`EDAM reduced operations`, "Other"))
```

### gather highlighting Galaxy wrapper tools for plotting

```{r}
df2 <- df2[order(-runs, -users)]
hg0 <- df2[1:15]
```

### create graph

```{r}
ncolors <- df2$cluster |> unique() |> length()

xvar <- xvar |> str_replace("Suite runs", "**Suite runs**")
yvar <- yvar |> str_replace("Suite users", "**Suite users**")

a <- df2 |>
    
    ggplot(aes(runs, users)) +
    
    geom_point(
        shape = 21, size = 2, stroke = .25, 
        aes(fill = cluster, color = cluster)
    ) +
    
    geom_text_repel(
        data = hg0, aes(label = `Suite ID`),
        bg.r = .05, bg.color = "grey96", fontface = "bold", family = "Calibri", box.padding = .5,
        segment.size = .3, max.overlaps = Inf, size = 3
    ) +
    
    scale_x_continuous(
        trans = "log10",  # expand = c(0, 0), 
        limits = c(1, 10000000), 
        breaks = scales::trans_breaks("log10", function(x) 10^x),
        labels = scales::trans_format("log10", scales::math_format(10^.x))
    ) +
    
    scale_y_continuous(
        trans = "log10", # limits = c(1, 10000),
        labels = scales::comma, # expand = c(0, 0), 
        breaks = c(.1, 1, 10, 100, 1000, 10000)
    ) +
    
    scale_fill_manual(values = c(paletteer_d("ggsci::default_igv", ncolors - 1), "grey") |> lighten(.25), guide = guide_legend(nrow = 4, override.aes = list(size = 2.5))) +
    scale_color_manual(values = c(paletteer_d("ggsci::default_igv", ncolors - 1), "grey") |> darken(.25), guide = "none") +
    
    coord_cartesian() +
    
    theme_minimal(base_family = "Calibri") +
    
    theme(
        legend.position = "bottom",
        legend.title.position = "top",
        legend.justification = "left",
        
        strip.text = element_markdown(),
        
        axis.title.x = element_markdown(margin = margin(t = 10)),
        axis.title.y = element_markdown(margin = margin(r = 10)),
        
        # axis.ticks = element_line(linewidth = .3),
        panel.grid.major = element_line(linewidth = .3, linetype = "dashed", lineend = "round", color = "grey75"),
        
        # panel.border = element_rect(linewidth = .3, fill = NA),
        
        axis.line = element_line(lineend = "round"),
        axis.ticks = element_line(lineend = "round")
    ) +
    
    labs(x = xvar, y = yvar, fill = "EDAM operation")
```

## No. of Tool Suites over time 

```{r}
d1 <- tools_dt[order(`Suite first commit date`)]

d1$ypos <- seq_along(d1$`Suite first commit date`)

d1$year <- d1$`Suite first commit date` |> lubridate::year()

d2 <- d1[, by = year, .(N = max(ypos))]

d2$year <- paste0(d2$year, "-01-01") |> lubridate::as_date()

d1_p <- d1 |>
    ggplot(aes(`Suite first commit date`, ypos)) +
    geom_line(linewidth = .45, color = "#3C3C3C") +
    geom_area(alpha = .25) +
    
    # geom_point(data = d2, aes(year, N)) +
    scale_x_date(expand = c(0.01, 0.01)) +
    scale_y_continuous(expand = c(0, 0), breaks = seq(50, 300, by = 50), limits = c(0, 300))+
    theme_minimal(base_family = "Calibri") +
    theme(
        # panel.grid.major = element_line(linewidth = .45, color = "grey85"),
        # panel.grid.minor = element_line(linewidth = .35),
        
        panel.grid.major = element_line(linewidth = .3, linetype = "dashed", lineend = "round", color = "grey75"),
        
        axis.title.x = element_markdown(margin = margin(t = 10)),
        axis.title.y = element_markdown(margin = margin(r = 10)),
        
        axis.line = element_line(lineend = "round"),
        axis.ticks = element_line(lineend = "round")
    ) +
    labs(y = "Cumulative number of tool suites")
```

## Patchwork 

```{r}

library(patchwork)

multi <- (free(c_1) | (a / d1_p)) + 
    plot_layout(widths = c(2, 1.5)) +
    plot_annotation(tag_levels = "A") &
    theme(
        plot.tag = element_text(face = "bold", family = "Calibri", size = 25),
        plot.background = element_rect(fill = "transparent", color = NA)
    )

ggsave(
    plot = multi, filename = "../static/extended-figures/extended_figure_3.svg",
    width = 14, height = 16, unit = "in"
)
```

::: {.callout-note}
Click on the Figure for pop up.
:::

![**Availability, usability, and growth of microbiology-related tool suites within the Galaxy ecosystem.** (A) Heatmap illustrating the availability of microbiology-related tool suites (x-axis) across various Galaxy servers (y-axis), grouped by EDAM topics. Tool suites may appear in multiple clusters, as they can be annotated with several topics. A logarithmic scale is applied for improved visualization. (B) Scatter plots showing the usage of microbiology-related tool suites over the past five years across all Galaxy main servers, categorized by their EDAM operations. The x-axis represents the total number of tool users, while the y-axis also denotes the total number of tool users (potentially correct this redundancy). (C) Cumulative number of tools added to Galaxy over time, based on the date of the first commit for each tool suite.](../static/extended-figures/extended_figure_3.svg)