---
title: "Tutorials"
lightbox: 
    match: auto
    effect: zoom
    desc-position: right
execute:
    eval: true
    warning: false
---

## Load `libraries`

```{r}
# install.packages("data.table")
library(data.table)
# install.packages("stringr")
library(stringr)

# install.packages(c("ggplot2", "ggrepel", "ggtext", "ggh4x"))
library(ggplot2)
library(ggrepel)
library(ggtext)
library(ggh4x)

library(shadowtext)

# install.packages("extrafont")
library(extrafont)

library(packcircles)
```

## Load input dataset

```{r}

## THIS SECTION MIGHT THROUGH AN ERROR
## THERE IS A BUG UNFORTUNATELY WITH fread FUNCTION
## RUN ON CONSOLE IN THIS CASE

microGalaxy_tutorials_url <- "https://raw.githubusercontent.com/usegalaxy-eu/microgalaxy_paper_2025/refs/heads/main/docs/supplementary/supplementary_table_4.tsv"
microGalaxy_tools_url     <- "https://raw.githubusercontent.com/usegalaxy-eu/microgalaxy_paper_2025/refs/heads/main/docs/supplementary/supplementary_table_2.tsv"
# all_tools_url             <- "https://raw.githubusercontent.com/galaxyproject/galaxy_codex/refs/heads/main/communities/all/resources/tools.tsv"


microGalaxy_tutorials_dt <- microGalaxy_tutorials_url |> fread()
microGalaxy_tools_dt     <- microGalaxy_tools_url |> fread()
# all_tools_dt            <- all_tools_url |> fread()

```

## Compute number of tools and number of microGalaxy tools per tutorial

```{r}

# 1 ------------------------

dt1 <- microGalaxy_tutorials_dt$Tools |> 
    str_split("\\,") |>
    lapply(str_squish) |>
    lapply(function(q) data.table("Tool ID" = q)) |>
    rbindlist(idcol = "Topic_id")

dt1$Topic <- microGalaxy_tutorials_dt[dt1$Topic_id]$Topic
dt1$Title <- microGalaxy_tutorials_dt[dt1$Topic_id]$Title

dt1$`EDAM topic`     <- microGalaxy_tutorials_dt[dt1$Topic_id]$`EDAM topic`
dt1$`EDAM operation` <- microGalaxy_tutorials_dt[dt1$Topic_id]$`EDAM operation`

tmp <- microGalaxy_tools_dt$`Tool IDs` |>
    str_split("\\,") |>
    lapply(str_squish) |>
    lapply(function(q) data.table("Tool ID" = str_squish(q))) |>
    rbindlist(idcol = "id")

tmp$`Suite ID` <- microGalaxy_tools_dt[tmp$id]$`Suite ID`

index <- match(dt1$`Tool ID`, tmp$`Tool ID`)

dt1$`Suite ID` = tmp[index]$`Suite ID`
```

### Exclude unnecessary columns

```{r}
microGalaxy_tutorials_dt$Topic                                      <- NULL
microGalaxy_tutorials_dt$Link                                       <- NULL
microGalaxy_tutorials_dt$`Servers with precise tool versions`       <- NULL
microGalaxy_tutorials_dt$`Servers with tool but different versions` <- NULL
```

## Tools coverage graph: Heatmap

```{r}

dt1$`Suite ID` <- ifelse(is.na(dt1$`Suite ID`) | dt1$`Suite ID` == "", "Not Available", dt1$`Suite ID`)

dt1 <- dt1[which(`Suite ID` != "Not Available")]

dt1$value <- 1

mm <- dt1[, c("Title", "Suite ID", "value"), with = FALSE] |> 
    unique() |> 
    dcast(Title ~ `Suite ID`, value.var = "value", fill = 0) |> 
    as.matrix(rownames = "Title")


hc_x <- mm |> t() |> dist(method = "binary") |> hclust("ward.D2")
hc_y <- mm |> dist(method = "binary") |> hclust("ward.D2")


# c_r <- dt1[, by = Title, .(N = `Suite ID` |> unique() |> length())]
# r_r <- dt1[, by = `Suite ID`, .(N = Title |> unique() |> length())]

# c_r <- c_r[order(N)]
# r_r <- r_r[order(-N)]

dt1$Title <- dt1$Title |> factor(hc_y$labels[hc_y$order] |> rev())
dt1$`Suite ID` <- dt1$`Suite ID` |> factor(hc_x$labels[hc_x$order] |> rev())

a_1 <- dt1 |>
    ggplot(aes(`Suite ID`, Title)) +
    geom_tile(color = "grey96") +
    
    scale_x_discrete(expand = c(0, 0)) +
    scale_y_discrete(expand = c(0, 0)) +
    
    theme_minimal(base_family = "Calibri") +
    theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        
        panel.border = element_rect(fill = NA, linewidth = .3),
        axis.ticks.x = element_line(linewidth = .3),
        
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(linetype = "dashed", lineend = "round", linewidth = .35),
        
        plot.margin = margin(10, 10, 10, 10)
    ) +
    
    labs(y = "Tutorials", x = "Galaxy Tool Suites")


edam_operations <- dt1$`EDAM operation` |>
    str_split("\\,") |>
    unlist() |>
    str_squish() |>
    table() |>
    as.data.table()

edam_topics <- dt1$`EDAM topic` |>
    str_split("\\,") |>
    unlist() |>
    str_squish() |>
    table() |>
    as.data.table()


edam_operations <- edam_operations[order(-N)] |> head(4)
edam_topics     <- edam_topics[order(-N)] |> head(4)


dt2 <- dt1 |> tidyr::separate_rows("EDAM topic", sep = ",") |> setDT()
dt2 <- dt2 |> tidyr::separate_rows("EDAM operation", sep = ",") |> setDT()

dt2$`EDAM topic`     <- dt2$`EDAM topic` |> str_squish() |> str_wrap(15)
dt2$`EDAM operation` <- dt2$`EDAM operation` |> str_squish() |> str_wrap(15)

edam_topics$V1     <- edam_topics$V1 |> str_wrap(15)
edam_operations$V1 <- edam_operations$V1 |> str_wrap(15)

dt2$`EDAM topic`     <- ifelse(dt2$`EDAM topic` %in% edam_topics$V1, dt2$`EDAM topic`, "Other")
dt2$`EDAM operation` <- ifelse(dt2$`EDAM operation` %in% edam_operations$V1, dt2$`EDAM operation`, "Other")

dt2$`EDAM topic`     <- dt2$`EDAM topic` |> factor(c(edam_topics$V1, "Other")) 
dt2$`EDAM operation` <- dt2$`EDAM operation` |> factor(c(edam_operations$V1, "Other")) 

a_2 <- dt2 |>
    ggplot(aes(`Suite ID`, Title)) +
    geom_tile(color = "white") +
    
    scale_x_discrete(expand = c(0, 0)) +
    scale_y_discrete(expand = c(0, 0)) +
    
    # facet_grid(rows = vars(edam_clean), scales = "free", space = "free") +
    
    # facet_nested_wrap(vars(`EDAM topic`, `EDAM operation`)) +
    
    facet_grid2(
        rows = vars(`EDAM topic`),
        cols = vars(`EDAM operation`),
        space = "free",
        scales = "free"
    ) +
    
    theme_minimal(base_family = "Calibri") +
    theme(
        axis.text.x = element_text(size = 4, angle = 45, hjust = 1),
        axis.text.y = element_text(size = 6),
        
        axis.title.x = element_text(margin = margin(t = 10)),
        axis.title.y = element_text(margin = margin(r = 10)),
        
        panel.border = element_rect(fill = NA, color = "grey", linewidth = .3),
        # axis.ticks = element_line(linewidth = .3, color = "grey"),
        
        strip.text.x = element_text(face = "bold"),
        strip.text.y = element_text(face = "bold", angle = 0, hjust = 0),
        
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(linetype = "dashed", lineend = "round", linewidth = .15),
        
        plot.margin = margin(10, 10, 10, 10)
    ) +
    
    labs(y = "Tutorials", x = "Galaxy Tool Suites")

ggsave(
    plot = a_2, filename = "../static/extended-figures/extended_figure_5.svg",
    width = 16, height = 10, unit = "in"
)
```

::: {.callout-note}
Click on the Figure for pop up.
:::

![**Usage of microbiology-related tool suites across training materials.** Heatmap illustrating the presence of microbiology-related tool suites within the available microbiology-related training materials, grouped by EDAM topics. The tool suites are organized based on their corresponding EDAM operations, highlighting the breadth of tools utilized for different training contexts.](../static/extended-figures/extended_figure_5.svg)

## EDAM ontology distribution: packcircles

```{r}

# ----------------------
d1 <- microGalaxy_tutorials_dt[, c("Title", "EDAM topic"), with = FALSE] |>
    tidyr::separate_rows("EDAM topic", sep = ",") |>
    setDT()

d2 <- microGalaxy_tutorials_dt[, c("Title", "EDAM operation"), with = FALSE] |>
    tidyr::separate_rows("EDAM operation", sep = ",") |>
    setDT()

# ---------------------

d1$`EDAM topic` = d1$`EDAM topic` |> str_squish()
d1$`EDAM topic` = ifelse(d1$`EDAM topic` == "", "No Topic", d1$`EDAM topic`)
    
d2$`EDAM operation` = d2$`EDAM operation` |> str_squish()
d2$`EDAM operation` = ifelse(d2$`EDAM operation` == "", "No Operation", d2$`EDAM operation`)

d1 = d1[, by = `EDAM topic`, .(N = Title |> unique() |> length())]
d2 = d2[, by = `EDAM operation`, .(N = Title |> unique() |> length())]

# ------------------------------

d1$`EDAM topic length`     <- d1$`EDAM topic` |> str_length()
d2$`EDAM operation length` <- d2$`EDAM operation` |> str_length()

d1 <- d1[order(-N, `EDAM topic`)]
d2 <- d2[order(-N, `EDAM operation`)]

# Generate the layout. 
# sizetype can be area or radius, 
# following your preference on what to be proportional to value.

packing_d1 <- circleProgressiveLayout(d1$N, sizetype = 'area') |> circleLayoutVertices(npoints = 100) |> setDT()
packing_d2 <- circleProgressiveLayout(d2$N, sizetype = 'area') |> circleLayoutVertices(npoints = 100) |> setDT()

packing_d1$strip <- d1[packing_d1$id]$`EDAM topic`
packing_d2$strip <- d2[packing_d2$id]$`EDAM operation`

packing_d1$`No. of Tutorials` <- d1[packing_d1$id]$N
packing_d2$`No. of Tutorials` <- d2[packing_d2$id]$N

packing_d1$pack <- "EDAM topic"
packing_d2$pack <- "EDAM operation"

d <- rbind(packing_d1, packing_d2)

h <- d[which(`No. of Tutorials` > 2), by = .(pack, strip, `No. of Tutorials`), .(
    x = (max(x) + min(x)) / 2,
    y = (max(y) + min(y)) / 2
)]

h$strip <- h$strip |> str_wrap(width = 8)

h$strip = h$strip |>
    str_replace_all("Phylogenetic\ntree", "Phylogenetic tree") |>
    str_replace_all("Box-Whisker\nplot", "Box-Whisker") |>
    str_replace_all("Public\nhealth", "Public health") |>
    str_replace_all("Nucleic\nacid", "Nucleic acid") |>
    str_replace_all("sequence\nanalysis", "sequence analysis")

b <- d |>
    ggplot(aes(x, y, group = id)) + 
    geom_polygon(aes(fill = `No. of Tutorials`), color = "grey96", linewidth = 1) +

    # geom_text_repel(
    #     data = h[which(`No. of Tutorials` < 15)], aes(x, y, label = strip, size = `No. of Tutorials`), 
    #     inherit.aes = FALSE, max.overlaps = Inf, bg.color = "grey96", color = "grey10",
    #     bg.r = .065, family = "Calibri"
    # ) +
        
    geom_shadowtext(
        data = h, aes(x, y, label = strip, size = `No. of Tutorials`),
        inherit.aes = FALSE, bg.color = "grey90", color = "grey20", bg.r = .065,
        family = "Calibri", fontface = "bold"
    ) +
    
    scale_fill_stepsn(
        colors = c('#00429d', '#2d58a7', '#456fb1', '#5986b9', '#6e9ec1', '#84b6c6', '#c6c6a9') |> rev(),
        guide = guide_colorsteps(
            barwidth = unit(10, "lines"),
            barheight = unit(.5, "lines")
        )
    ) +
    
    scale_size_continuous(range = c(3, 4), guide = "none") +
    
    facet_wrap(vars(pack), nrow = 1) +
    theme_void(base_family = "Calibri") + 
    theme(
        legend.position = "bottom",
        legend.title.position = "top",
        
        strip.text = element_text(face = "bold")
    ) +
    coord_equal()

```